Boosting Higgs Pair Production in the $bb\bar{b}\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-06 paper/sec-mva.tex
MVA

The final step of our analysis is to process our three exclusive
categories of events (boosted, intermediate and resolved) though
a Multivariate Analysis.
%
The particular tool that we adopt is
a feed-forward artificial neural network.\footnote{This is the
  same type of neural networks as those used in the
  NNPDF analysis of parton distributions~\cite{Ball:2010de}.}
%
The architecture of the neural network is
\be
\label{eq:nn1}
N_{\mathrm{var}}\times5\times3\times1 \, ,
\ee
with $N_{\mathrm{par}}$ the number of input variables for the MVA.
%
Therefore, we have two hidden layers and a single output variable, which
is the value of the discriminant to separate signal and background.\footnote{Note that in the MVA literature artificial neural networks with several hidden
  layers are known as {\it deep neural networks}.}
%
Below we list the variables that are given as input to the MVA in each
different category.
%
The MVA has trained using a Genetic Algorithm during 50K iterations,
and the figure of merit to be minimized is the cross-entropy function.
%
This figure of merit is defined as
\be
\bf DEF
\ee

For both signal and background events, the variables that we input
to the MVA are the following.
%
For the boosted category we have
\begin{itemize}
\item The invariant mass of the two mass-drop-tagged jets, $m_{\rm jet}$
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$
\item The rapidity of the two mass-drop-tagged jets, $y_{\rm jet}$
\item The invariant mass of the leading Higgs candidate, $m_{h,1}$
\item The invariant mass of the sub-leading Higgs candidate, $m_{h,2}$
\item The splitting scale of the first Higgs
\item The splitting scale of the second Higgs
\item The subjetiness variable $\tau_{21,h1}$ of the leading Higgs candidate
\item The subjetiness variable $\tau_{21,h2}$
  of the sub-leading Higgs candidate
 \item The angular separation between the two tagged jets, $\Delta R_{1,2}$ 
\end{itemize}
For completeness, we give here the definition of the various substructure variables.
%
The splitting scale is defined as
\be
\bf DEF
\ee
while the $\tau_{1,2}$ subjetiness variable is defined as
\be
\bf DEF
\ee
The other variables are more conventional.
%
As will be shown below, the addition of the jet substructure variables,
the splitting scales and the subjetiness,
enhances substantially the signal significance, as compared
to using only jet variables without substructure information.

For the resolved analysis instead, the $N_{\rm var}$ kinematic variables
that are given as input to the MVA are the following
\begin{itemize}
\item The invariant mass of the four $b$-tagged leading jets (that
  is, of the di-Higgs system), $m_{4b}$
\item The transverse momentum of the di-Higgs system, $p_{4b}$
\item The rapidity of the di-Higgs system $y_{4b}$
\item The invariant mass of the leading Higgs candidate $m_{h,1}$
\item The invariant mass of the subleading Higgs candidate $m_{h,1}$
\item The angular separation between each of the four
  $b$-tagged leading jets in the event: $\Delta R_{b1,b2}$,
  $\Delta R_{b1,b3}$, $\Delta R_{b1,b4}$, $\Delta R_{b2,b3}$,
  $\Delta R_{b2,b4}$, $\Delta R_{b3,b4}$
\end{itemize}
As will be shown below, of all these variables, the invariant mass of the
Higgs candidates is the one that carries as higher discrimination
power.

\subsection{Enhancing signal discrimination with the MVA}

>>>>

The classification of events as arising from either signal or
background processess via the use of a multivariate analysis is a common
technique in high energy physics.

In order to investigate the potential efficacy of an MVA in an $HH\to 4b$
measurement, we process the events remaining after the cut-based analysis 
through a standard feed-forward artificial neural network with architecture

\be
\label{eq:nn1}
N_{\mathrm{var}}\times5\times3\times1 \, ,
\ee
where $N_{\mathrm{par}}$ represents the number of input variables for the MVA.
All neural-network layers use a sigmoid activation function, allowing for a probabilistic
interpretation of the neural-network output.

The input variables for the MVA differ between our three categories, based upon
the availability in each analysis of substructure information on fat jets. For all
three analyses, the following common variables are used:

\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs candidates, $p_{T,H_0}$, $p_{T,H_1}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,HH}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{H,1}$, $m_{H,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{HH}$.
\item The separation in $R$ of the two Higgs candidates, $\Delta R_{HH}$.
\item The separation in $\phi$ of the two Higgs candidates, $\Delta \phi_{HH}$.
\item The separation in $\eta$ of the two Higgs candidates, $\Delta \eta_{HH}$.
\item The transverse momenta of the leading Higgs candidate subjets, $p_{T,H_{0,0}}$, $p_{T,H_{0,1}}$
\item The transverse momenta of the sub-leading Higgs candidate subjets, $p_{T,H_{1,0}}$, $p_{T,H_{1,1}}$
\end{itemize}

For analyses including the reconstruction of fat jets, the following variables are included for
every fat jet Higgs candidate present in the analysis.

\begin{itemize}
\item The $k_T$ splitting scale of the fat-jet (Reference).
\item The ratio of 2-to-1-Subjettiness $\tau_{2}/\tau_{1}$ of the fat-jet (Reference).
\item The fat-jet energy correlation function double-ratio $C_2$ (Reference).
\item The fat-jet energy correlation function double-ratio $D_2$ (Reference).
\end{itemize}

Therefore there are in total $N_{\mathrm{var}}=13$ variables for the resolved analysis, 17 for the semi-resolved, and 21 for the boosted category.

Given the above set of kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network parameters $\{\omega\}$, we interpret the neural network output $y_i$ as the probability that the event $i$ originates from the signal process,
\be y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} ), \ee
where $y_i^\prime$ represents the true classification of the event $i$, i.e $y^\prime_{\text{signal}} = 1$ and $y^\prime_{\text{background}} = 0$. With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i}.
\ee
 Correspondingly the negative log-liklihood cost function to be used in the neural network training is the cross-entropy error
\be E(\{\omega\}) = -\log\left(\prod_i^{N_{\text{evt}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right) = \sum_i^{N_{\text{evt}}} y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)}. \ee

Using the cross-entropy cost function, the neural networks are trained with a basic genetic algorithm (GA) over 50,000 generations.
(Do we need to specifiy GA params etc?).
\subsection{Enhancing signal discrimination with the MVA}
